---
title: "Breast Exploratory Analysis"
author: "Jonathan Ma"
date: "`r Sys.Date()`"
output:
  word_document:
    toc: true
    number_sections: true
---

\newpage

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, message=FALSE, warning=FALSE, fig.width=6, fig.height=4)
library(tidyverse)
library(dplyr)
library(ggplot2)
library(stringr)
library(ggthemes)
library(scales)
```

# Introduction

This exploratory data analysis (EDA) is the first step in our project pipeline:

1. Bayesian Hierarchical Modeling
2. Posterior Inference & Diagnostics
3. Clustering of Individual Risk Profiles

We use data from the [SEER Program](https://seer.cancer.gov/) focused on breast cancer cases, structured across three levels:

- **Level 1 (Tumor):** Tumor-specific details (e.g., stage, size, grade)
- **Level 2 (Patient):** Demographics and marital status
- **Level 3 (Region):** County-level metro/rural designations

# Data Structure and Levels

| Level    | Variable | Description                                    |
|----------|----------|------------------------------------------------|
| Tumor    | stage    | SEER stage: Localized, Regional, Distant       |
| Tumor    | grade    | Tumor grade: 1–4 or unknown                    |
| Tumor    | size     | EOD Tumor Size (mm)                            |
| Tumor    | site     | Tumor site (ICD code)                          |
| Patient  | id       | Patient identifier                             |
| Patient  | age      | Age group at diagnosis                         |
| Patient  | sex      | Biological sex                                 |
| Patient  | raceth   | Race/ethnicity                                 |
| Patient  | year     | Year of diagnosis                              |
| Patient  | marry    | Marital status                                 |
| Region   | region   | County-level metro/rural code                  |


```{r Data Loading}
seer <- read.delim("../data/raw/breastcancer.txt", sep = "\t", header = FALSE, stringsAsFactors = FALSE)
```

We replace "Unknown" and "Blank(s)" with `NA`, and then analyse missing data. We want to see if missing data is specific to a group so we can determine if the missingness is MCAR or not.

```{r Rename}
seer_prime <- seer %>%
  transmute(
    id     = V1,
    region = V2,
    stage  = V3,
    age    = V4,
    sex    = V5,
    raceth = V6,
    grade  = V7,
    size   = V10,
    year   = V11,
    marry  = V12,
    site = V8
  )

seer_prime <- seer_prime %>%
  mutate(across(where(is.character), ~na_if(., "Unknown"))) %>%
  mutate(across(where(is.character), ~na_if(., "Blank(s)")))
```

Is missing data more common in older patients?

```{r}
df <- seer_prime

df <- seer_prime %>%
  mutate(size_missing = is.na(size))

ggplot(df, aes(x = age, fill = size_missing)) +
  geom_bar(position = "fill") +
  labs(title = "Proportion of Missing Tumor Size by Age",
       y = "Proportion Missing",
       x = "Age Group",
       fill = "Size Missing") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Is missing data associated with Race/Ethnicity?

```{r}
table(df$raceth, is.na(df$size))
```

Is stage more likely to be missing in nonmetro rural areas?

```{r}
df <- df %>%
  mutate(grade_missing = is.na(grade),
         region_type = factor(region))

# Proportions
prop.table(table(df$region, df$grade_missing), 1)
```

Although we did not do a statistical test for MCAR, visual and tabular inspection clearly show patterns in missingness. Tumor size by age is non-uniform, with younger and older ends having higher missing rates. Missing rates also vary by racial group, with whites having notably higher missingness.
 
Logistic regression is a standard method for exploring missing data mechanisms. By modeling the missingness indicator (e.g., stage_miss) as a binary outcome and regressing it on observed variables, we can test whether missingness is related to any known covariates.

If none of the predictors are significant, this supports the MCAR assumption. But if missingness is significantly associated with observed variables (as seen here), MCAR is rejected, and MAR becomes a more reasonable assumption for downstream modeling.

```{r}
library(broom)
# Create a missingness indicator for tumor size
df$stage_miss <- is.na(df$stage)

# Convert to factor if not already
df$sex <- factor(df$sex)
df$raceth <- factor(df$raceth)
df$region <- factor(df$region)
df$age <- factor(df$age)
df$marry <- factor(df$marry)
df$grade <- factor(df$grade)
df$size <- factor(df$size)

# Fit logistic regression model to predict missingness
miss_model <- glm(stage_miss ~ sex + raceth + region + age + marry + grade + size,
                  data = df, family = binomial)

# Load required package
library(broom)

# Tidy the model output, add significance stars, and filter for p < 0.05
miss_summary <- broom::tidy(miss_model) |>
  dplyr::filter(p.value < 0.05) |>
  dplyr::mutate(
    signif = dplyr::case_when(
      p.value < 0.001 ~ "***",
      p.value < 0.01  ~ "**",
      p.value < 0.05  ~ "*",
      TRUE ~ ""
    )
  ) |>
  dplyr::select(term, signif)

# View the cleaned summary
print(miss_summary)
```

The missingness in stage is likely not MCAR, because missingness depends significantly on several observed variables (e.g., sex, raceth, region, age, marry, grade, size). Therefore, it’s reasonable to assume MAR (Missing at Random) for modeling purposes. We find we have 209123 cases to work with (4.06% of the whole thing) after removing "Unknown" and "Blank(s)" Values.

```{r}
seer_prime <- seer_prime %>%
  filter(complete.cases(.))

total_cases <- nrow(seer)
complete_cases <- nrow(seer_prime)
percent_complete <- round(100 * complete_cases / total_cases, 2)

cat("Complete cases:", complete_cases, "out of", total_cases,
    sprintf("(%s%% complete)\n", percent_complete))
```

# Variable-Specific Treatment

Further cleaning is done via tabling all variables. 

We want to filter for top ten cancer sites (breast, lung, etc). We exclude 998 and 999 because they are unknown.

```{r SITE}
seer_prime <- seer_prime %>%
  filter(!(site %in% c("998", "999")))

top_sites <- seer_prime %>%
  count(site, sort = TRUE) %>%
  slice_head(n = 10)

seer_prime <- seer_prime %>%
  filter(site %in% top_sites$site)

table(seer_prime$site)
```

Collapse and recode into ordered, interpretable regional factors, removes the Alaska/Hawaii “unknown” entries.

```{r REGION}
seer_prime <- seer_prime %>%
  mutate(
    region = case_when(
      str_detect(region, "1 million") ~ "Large",
      str_detect(region, "lt 250") | str_detect(region, "250,000") ~ "Small",
      str_detect(region, "adjacent") | str_detect(region, "not adjacent") ~ "Nonmetro",
      TRUE ~ NA_character_
    ),
    region = factor(region, levels = c("Large", "Small", "Nonmetro"))
  )
table(seer_prime$region)
```

Restrict to tumors that are invasive (drop In situ, Unknown/unstaged).

```{r STAGE}
seer_prime <- seer_prime %>%
  mutate(
    stage = case_when(
      stage %in% c("Localized", "Regional", "Distant") ~ stage,
      TRUE ~ NA_character_
    ),
    stage = factor(stage, levels = c("Localized", "Regional", "Distant"))
  )
table(seer_prime$stage)
```

Extract numeric grade 1–4; drop cell‑line entries (“B‑cell”, “T‑cell” etc.). Convert them to Start (1/2) and End (3/4) since our model is logistic.

```{r GRADE}
seer_prime <- seer_prime %>%
  mutate(
    grade = case_when(
      str_detect(grade, "I$") ~ "1",
      str_detect(grade, "II$") ~ "2",
      str_detect(grade, "III$") ~ "3",
      str_detect(grade, "IV$") ~ "4",
      TRUE ~ NA_character_
    ),
    grade = case_when(
      grade %in% c("1", "2") ~ "Start",
      grade %in% c("3", "4") ~ "End",
      TRUE ~ NA_character_
    ),
    grade = factor(grade, levels = c("Start", "End"), ordered = FALSE)
  )
table(seer_prime$grade)
```

Making this numeric and then filtering implausible values. 

```{r SIZE}
seer_prime <- seer_prime %>%
  mutate(
    size = as.numeric(size),
    size = ifelse(size > 0 & size <= 200, size, NA)
  )
summary(seer_prime$size)
```

Making this binary, only married or unmarried. 

```{r MARRY}
seer_prime <- seer_prime %>%
  mutate(
    marry = case_when(
      str_detect(marry, "Married") ~ "Married",
      TRUE ~ "Unmarried"
    ),
    marry = factor(marry)
  )
table(seer_prime$marry)
```

Recode RACETH to be three levels, Black, White, and Other.

```{r RACISM}
seer_prime <- seer_prime %>%
  mutate(
    raceth = case_when(
      raceth == "White" ~ "W",
      raceth == "Black" ~ "B",
      TRUE ~ "O"  
    ),
    raceth = factor(raceth, levels = c("W", "B", "O"))
  )
table(seer_prime$raceth)
```

## Complete Summary of available data

Seeing how many cases are available now:

```{r Checking}
seer_prime <- seer_prime %>% filter(complete.cases(id, region, stage, age, sex, raceth, grade, size, site, marry, year))
cat("Remaining complete cases:", nrow(seer_prime), "\n")
```

# Hierarchal Encoding

To do hierarchal models in R, identifiers must be encoded as numeric integers (e.g., 1, 2, ..., N) for levels like REGION and ID. We complete that, then set factor variables, and check resulting data structure.

```{r Hierarchy prep}
seer_prime <- seer_prime %>%
  mutate(
    regionid = as.integer(factor(region)),
    patientid = as.integer(factor(id))
  )

seer_prime <- seer_prime %>%
  mutate(
    age = factor(age),
    sex = factor(sex),
    raceth = factor(raceth),
  )

str(seer_prime)

# write.csv(seer_prime, "../data/clean/seer_nov13.csv", row.names = FALSE)
```