---
title: "Posterior Summary"
author: "Jonathan Ma"
date: "`r Sys.Date()`"
output:
  word_document: default
  html_document:
    df_print: paged
---

```{r setup, include=FALSE, messages=FALSE}
knitr::opts_chunk$set(echo = TRUE)
brm_model2 <- readRDS("../R/breast_model.rds")
library(dplyr)
library(ggplot2)
df <- read.csv("../data/clean/seer_df2.csv", sep = ",", header = TRUE, stringsAsFactors = TRUE)

```

# Posterior Summary


## Summary 

The following posterior summary, random and fixed effects summary code is provided, when we cluster we can save these as dataframes and move from there. 


```{r Posterior Summary, echo=TRUE}
posterior_summary <- posterior_summary(brm_model2)
```

## Fixed Effects

```{r Fixed Effects, echo=TRUE}
fe <- fixef(brm_model2, probs = c(0.025, 0.975))
```

## Random Effects (group level variation)

```{r Random Effects, echo=TRUE}
re <- ranef(brm_model2)
```

## Odds Ratios and Marginal Effects

The odds ratio (OR) is the exponentiated posterior mean (or median) of each coefficient.

```{r OR, echo=TRUE}
odds_ratios <- exp(fe)
odds_df <- as.data.frame(odds_ratios)
odds_df$variable <- rownames(odds_df)
odds_df <- odds_df[order(abs(odds_df$Estimate - 1), decreasing = TRUE), ]
head(odds_df, 10)
```

```{r Marginal Effects, echo=TRUE}
me_list <- marginal_effects(brm_model2)
me_combined <- bind_rows(lapply(names(me_list), function(v) {
  df <- me_list[[v]]
  df$variable <- v
  df
}))
me_combined <- me_combined %>%
  mutate(range = upper__ - lower__) %>%
  arrange(desc(range))
head(me_combined, 10)
```

## Shrinkage EFfects

```{r Shrinkage Effects, echo=TRUE}
patient_re <- as.data.frame(re$patientid)
region_re <- as.data.frame(re$regionid)
patient_re$group <- rownames(re$patientid)
region_re$group <- rownames(re$regionid)

# patient-level shrinkage
top_patient <- patient_re %>%
  arrange(desc(abs(Estimate.Intercept))) %>%
  slice(1:30)  # top 30 most extreme patients

ggplot(top_patient, aes(x = reorder(group, Estimate.Intercept), 
                        y = Estimate.Intercept)) +
  geom_point() +
  geom_errorbar(aes(ymin = Q2.5.Intercept, ymax = Q97.5.Intercept), width = 0.2) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray") +
  coord_flip() +
  labs(
    title = "Shrinkage of Patient-Level Intercepts",
    x = "Patient ID",
    y = "Estimated Intercept (log-odds scale)"
  ) +
  theme_minimal()

# region-level shrinkage
top_region <- region_re %>%
  arrange(desc(abs(Estimate.Intercept)))

ggplot(top_region, aes(x = reorder(group, Estimate.Intercept), 
                        y = Estimate.Intercept)) +
  geom_point() +
  geom_errorbar(aes(ymin = Q2.5.Intercept, ymax = Q97.5.Intercept), width = 0.2) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray") +
  coord_flip() +
  labs(
    title = "Shrinkage of Region-Level Intercepts",
    x = "Region ID",
    y = "Estimated Intercept (log-odds scale)"
  ) +
  theme_minimal()
```

### Raw vs Shrunk Estimates

```{r Comparison}
# Raw (non-pooled) observed rate of latestage by region
raw_region <- seer_df2 %>%
  group_by(regionid) %>%
  summarise(
    raw_rate = mean(latestage),
    n = n()
  )

# Extract region-level random intercepts (log-odds scale)
region_re <- as.data.frame(ranef(brm_model2)$regionid)
region_re$regionid <- rownames(ranef(brm_model2)$regionid)

# Rename for clarity
region_post <- region_re %>%
  select(regionid, estimate = Estimate.Intercept,
         lower = Q2.5.Intercept, upper = Q97.5.Intercept)

# Merge raw rates and posterior estimates
region_compare <- left_join(raw_region, region_post, by = "regionid") %>%
  mutate(
    posterior_prob = plogis(estimate),  # convert log-odds to probability
    lower_prob = plogis(lower),
    upper_prob = plogis(upper)
  )

library(ggplot2)

ggplot(region_compare, aes(x = reorder(regionid, raw_rate))) +
  geom_point(aes(y = raw_rate), color = "blue", size = 2, alpha = 0.6) +
  geom_point(aes(y = posterior_prob), color = "red", size = 2, alpha = 0.6) +
  geom_errorbar(aes(ymin = lower_prob, ymax = upper_prob), color = "red", width = 0.2, alpha = 0.5) +
  coord_flip() +
  labs(
    title = "Shrinkage: Raw vs Posterior Region-Level Late-Stage Probabilities",
    x = "Region ID",
    y = "Probability of Late-Stage Diagnosis",
    caption = "Blue = raw rate, Red = posterior mean ± 95% CI"
  ) +
  theme_minimal()
```

## Intra-Class Correlation

```{r ICC, echo=TRUE}
# Extract posterior draws of group-level standard deviations
posterior <- as_draws_df(brm_model2)

# Names like sd_patientid__Intercept, sd_regionid__Intercept
head(select(posterior, starts_with("sd_")))

# Compute posterior ICCs
posterior <- posterior %>%
  mutate(
    ICC_patientid = (sd_patientid__Intercept^2) / 
                    (sd_patientid__Intercept^2 + sd_regionid__Intercept^2 + pi^2 / 3),
    
    ICC_regionid = (sd_regionid__Intercept^2) / 
                   (sd_patientid__Intercept^2 + sd_regionid__Intercept^2 + pi^2 / 3)
  )

posterior %>%
  summarise(
    ICC_patientid_mean = mean(ICC_patientid),
    ICC_patientid_low  = quantile(ICC_patientid, 0.025),
    ICC_patientid_high = quantile(ICC_patientid, 0.975),
    
    ICC_regionid_mean = mean(ICC_regionid),
    ICC_regionid_low  = quantile(ICC_regionid, 0.025),
    ICC_regionid_high = quantile(ICC_regionid, 0.975)
  )
```

```{r ICC Density, echo=TRUE}
posterior %>%
  select(ICC_patientid, ICC_regionid) %>%
  pivot_longer(everything(), names_to = "Level", values_to = "ICC") %>%
  ggplot(aes(x = ICC, fill = Level)) +
  geom_density(alpha = 0.5) +
  labs(title = "Posterior ICC Distributions",
       x = "ICC", y = "Density") +
  theme_minimal()
```

## Posterior Plots

```{r PvP Plot, echo=TRUE}

```